### 2010-03-29 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: get some sort of introduction.
Desc@ICFP: Tidied up the code in the paper, fixed a few transcription bugs and added colour (well, purple)

\section{A universe of simple data-types}

@@

In dependently-typed language, $\Sigma$-types can be interpreted in
two different ways. This duality is actually reflected in the notation
we can find in the literature, depending on the sensibility of the author.
The notation $\Sigma_{x : A} (B\: x)$ stresses the interpretation that
$\Sigma$-types are a generalisation of sums over arbitrary arities.  Where
simply-typed languages have finite sums, dependently-typed languages have
sums indexed by any set.

On the other hand, our choice of notation ($\SIGMA{\V{x}}{A} (B\: x)$)
emphasises that $\Sigma$-types can be read as a generalisation of products,
where the second component can depend on the first one. When simply-typed
languages pack data into non-dependent tuples, dependently-types languages
tolerate this non-dependent usage. However, they also give the rather novel
ability for data to influence types of further data.

@@

We propose to embed inductive types as a universe in our dependent
type theory, the universe of \emph{descriptions}
(Figure~\ref{fig:desc_universe}). As expected, the code of this
universe mimics the standard grammar of data-types definitions in
simply-typed languages. Hence, we have a $\DSigma{}{}$ code,
interpreted as a $\Sigma$-type, to build the
sigmas-of-sigmas. Descriptions are terminated by $\DUnit$, which
contains no useful payload. The functoriality of the data-types is
introduced by $\DIndx$. When we tie the knot with a fix-point, the
hole left open by $\DIndx$ will be turned into a standard recursive
argument. We notice that this functoriality appears in the type of
$\descop{D}{}$ itself, for a given data-type definition $D$. This
corresponds to the object part of the functor described by $D$.

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DUnit          & \Desc \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
    \DIndx          & \Desc \To \Desc
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}ll@{\:=\:\:}ll}
\descop{\DUnit}{& X}        &  \Unit                                       \\
\descop{\DSigma{S}{D}}{& X} &  \SIGMAS{\V{s} : S}{\descop{D~s}{X}}         \\
\descop{\DIndx{D}}{& X}     &  \TIMES{X}{\descop{D}{X}}
\end{array}
}
\end{array}
}\]

@@

\[    \NatD\: Z \mapsto 1 + Z    \]

@@

\[\stk{
\ListD : \Set \To \Desc \\
\ListD \: X \mapsto \DSigma{(\EnumT{[ \ListNil, \ListCons ]})}
                           {[ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{\DUnit}}) ]}
}\]

The modification consists in turning the $\NatSuc{}$ constructor into
a proper $\ListCons$ taking an argument in X followed by a recursive
argument. In this case, we use $\DSigma{}{}$ in its product
interpretation: we pack an element of $X$ together with the recursive
argument. Easily, one sees that this code actually describes the
following functor:

\[    \ListD\: X\: Z \mapsto 1 + X \times Z     \]

@@

\[\stk{
\TreeD : \Set \To \Desc \\
\begin{array}{@{}ll}
\TreeD \: X \mapsto \DSigma{ & (\EnumT{[ \TreeLeaf, \TreeNode ]}) \\}
                           { & [ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{(\DIndx{\DUnit})}}) ]}
\end{array}
}\]

@@

\[    \TreeD\: X\: Z \mapsto 1 + X \times Z \times Z     \]

@@

Again, this is a clear reminiscence of the sum-of-product style:
we have a finite sum of constructors. Of course, every description can
be expressed in this style, using a singleton as single constructor if
necessary. A tagged description can trivially be turned into a
full-blown description. We note $\toDesc{TD}$ the description computed
from the tagged description $TD$.

@@

\[
\Rule{\Gamma \vdash \Bhab{D}{\Desc}}
     {\Gamma \vdash \Bhab{\Mu{D}}{\Set}} \qquad
\Rule{\Gamma \vdash \Bhab{D}{\Desc} \quad 
      \Gamma \vdash \Bhab{x}{\descop{D}{(\Mu{D})}}}
     {\Gamma \vdash \Bhab{\Con{x}}{\Mu{D}}}
\]

@@

\[
\F{cata} : \PITEL{D}{\Desc}
           \PI{T}{\Set}
           (\descop{D}{T} \To T) \To 
           \Mu{D} \To T 
\]

@@

The first step consists in turning $T$ into a predicate
$\Bhab{P}{\Mu{D} \To \Set}$, hence introducing a dependency on the
descriptions. Consequently, we strengthen the algebra $\descop{D}{T}
\To T$ by a similar process. We start with turning the arrow into a
dependent arrow, hence offering $\Bhab{xs}{\descop{D}{(\Mu{D})}}$ to
further analysis.

Then, we have to translate the notion of algebra in our dependent
setting. This is achieved by the predicate $\All{D}{X}{P}{x}$
(Fig.~\ref{fig:all-predicates}) that, intuitively, states that
$\Bhab{P}{X \To \Set}$ holds everywhere in $x$. Based on that
knowledge, we ought to be able to prove that $P$ holds for
$\Con{xs}$. We close our elimination principle by taking an argument
$x$ in $\Mu{D}$ and claiming that the motive $P$ is realised on $x$.

\[
\begin{array}{lcll}
\F{induction} & : & \multicolumn{2}{l}{\PITEL{D}{\Desc}}                   \\
              &   & \multicolumn{2}{l}{\PITEL{P}{\Mu{D} \To \Set}}         \\
              &   & \PITEL{m}{& \PI{xs}{\descop{D}{(\Mu{D})}}              \\
              &   &           & \All{D}{(\Mu{D})}{P}{xs} \To P (\Con{xs})} \\
              &   & \multicolumn{2}{l}{\PI{x}{\Mu{D}} P x}
\end{array}
\]

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\All{}{}{}{} : & \PITEL{\V{D}}{\Desc}
                 \PITEL{\V{X}}{\Set}
                 \PITEL{\V{P}}{\V{X} \To \Set} \\
               & \PI{\V{xs}}{\descop{D}{\V{X}}} 
                 \Set 
\end{array} \\
\All{\DId}{X}{P}{x} \mapsto P\: x \\
\All{(\DConst{Z})}{X}{P}{x} \mapsto Z \\
\All{(\DProd{D}{D'})}{X}{P}{\pair{d}{d'}{}} \mapsto \TIMES{\All{D}{X}{P}{d}}{\All{D'}{X}{P}{d'}} \\
\All{(\DSigma{S}{T})}{X}{P}{\pair{a}{b}{}} \mapsto \All{(T\: a)}{X}{P}{b} \\
\All{(\DPi{S}{T})}{X}{P}{f} \mapsto \PI{\V{s}}{S} \All{(T\: a)}{X}{P}{(f\: s)}
}
&
%%
\stk{
\begin{array}{@{}ll}
\all{}{}{}{}{} : & \PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X}}{\Set}
                   \PITEL{\V{P}}{\V{X} \To \Set} \\
                 & \PITEL{\V{R}}{\PI{\V{x}}{X} P\: x}
                   \PI{\V{xs}}{\descop{D}{\V{X}}} 
                   \All{D}{X}{P}{xs} 
\end{array} \\
\all{\DId}{X}{P}{R}{x} = R x \\
\all{(\DConst{Z})}{X}{P}{R}{x} = z \\
\all{(\DProd{D}{D'})}{X}{P}{R}{\pair{d}{d'}{}} = \pair{\all{D}{X}{P}{R}{d}}{\all{D'}{X}{P}{R}{d'}}{} \\
\all{(\DSigma{S}{T})}{X}{P}{R}{\pair{a}{b}{}} = \all{(T\: a)}{X}{P}{R}{b} \\
\all{(\DPi{S}{T})}{X}{P}{R}{f} = \LAM{s} \all{(T\: s)}{X}{P}{R}{(f\: s)}
\end{array}
}
\]

@@

\[
\Rule{\forall e_i \in \vec{e}, \propag{\push{e_i}
                                            {\descop{f\: i}{(\Mu{(\DSigma{(\Enum\: b)}{f})})}}}
                                      {e'_i}}
     {\propag{\push{C\: \vec{e}}{\Mu{(\DSigma{(\Enum\: b)}{f})}}}
             {\Con{(\etag{C}, \widetilde{e'})}}}\;\etag{C} \in b
\]

>>>>

\section{A universe of simple data-types}
\label{sec:universe-desc}

@@

In dependently-typed language, $\Sigma$-types can be interpreted in
two different ways. This duality is actually reflected in the notation
we can find in the literature, depending on the sensibility of the author.
The notation $\blue{\Sigma}_{\V{x} : \V{A}} (\V{B}\: \V{x})$ stresses the interpretation that
$\Sigma$-types are a generalisation of sums over arbitrary arities.  Where
simply-typed languages have finite sums, dependently-typed languages have
sums indexed by any set.

On the other hand, our choice of notation $\SIGMA{\V{x}}{\V{A}} (\V{B}\:\V{x})$
emphasises that $\Sigma$-types can be read as a generalisation of products,
where the second component can depend on the first one. When simply-typed
languages pack data into non-dependent tuples, dependently-types languages
tolerate this non-dependent usage. However, they also give the rather novel
ability for data to influence types of further data.

@@

We propose to embed inductive types as a universe in our dependent
type theory, the universe of \emph{descriptions}
(Figure~\ref{fig:desc_universe}). As expected, the code of this
universe mimics the standard grammar of data-types definitions in
simply-typed languages. Hence, we have a $\DSigma{}{}$ code,
interpreted as a $\Sigma$-type, to build the
sigmas-of-sigmas. Descriptions are terminated by $\DUnit$, which
contains no useful payload. The functoriality of the data-types is
introduced by $\DIndx$. When we tie the knot with a fix-point, the
hole left open by $\DIndx$ will be turned into a standard recursive
argument. We notice that this functoriality appears in the type of
$\descop{\V{D}}{}$ itself, for a given data-type definition $\V{D}$. This
corresponds to the object part of the functor described by $\V{D}$.

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DUnit          & \Desc \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
    \DIndx          & \Desc \To \Desc
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}ll@{\:=\:\:}ll}
\descop{\DUnit}{& \V{X}}        &  \Unit                                       \\
\descop{\DSigma{\V{S}}{\V{D}}}{& \V{X}} &  \SIGMAS{\V{s} : \V{S}}{\descop{\V{D}~\V{s}}{\V{X}}}         \\
\descop{\DIndx{\V{D}}}{& \V{X}}     &  \TIMES{\V{X}}{\descop{\V{D}}{\V{X}}}
\end{array}
}
\end{array}
}\]

@@

\[    \NatD\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{Z}    \]

@@

\[\stk{
\ListD : \Set \To \Desc \\
\ListD \: \V{X} \mapsto \DSigma{(\EnumT{[ \ListNil, \ListCons ]})}
                           {[ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{\DUnit}}) ]}
}\]

The modification consists in turning the $\NatSuc{}$ constructor into
a proper $\ListCons$ taking an argument in $\V{X}$ followed by a recursive
argument. In this case, we use $\DSigma{}{}$ in its product
interpretation: we pack an element of $\V{X}$ together with the recursive
argument. Easily, one sees that this code actually describes the
following functor:

\[    \ListD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{X} \D{\ensuremath{\times}} \V{Z}     \]

@@

\[\stk{
\TreeD : \Set \To \Desc \\
\begin{array}{@{}ll}
\TreeD \: \V{X} \mapsto \DSigma{ & (\EnumT{[ \TreeLeaf, \TreeNode ]}) \\}
                           { & [ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{(\DIndx{\DUnit})}}) ]}
\end{array}
}\]

@@

\[    \TreeD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{X} \Prod \V{Z} \Prod \V{Z}     \]

@@

Again, this is a clear reminiscence of the sum-of-product style:
we have a finite sum of constructors. Of course, every description can
be expressed in this style, using a singleton as single constructor if
necessary. A tagged description can trivially be turned into a
full-blown description. We write $\toDesc{\V{TD}}$ for the description computed
from the tagged description $\V{TD}$.

@@

\[
\Rule{\Gamma \vdash \Bhab{\V{D}}{\Desc}}
     {\Gamma \vdash \Bhab{\Mu{\V{D}}}{\Set}} \qquad
\Rule{\Gamma \vdash \Bhab{\V{D}}{\Desc} \quad 
      \Gamma \vdash \Bhab{\V{x}}{\descop{\V{D}}{(\Mu{\V{D}})}}}
     {\Gamma \vdash \Bhab{\Con{\V{x}}}{\Mu{\V{D}}}}
\]

@@

\[
\F{cata} : \PITEL{\V{D}}{\Desc}
           \PI{\V{T}}{\Set}
           (\descop{\V{D}}{\V{T}} \To \V{T}) \To 
           \Mu{\V{D}} \To \V{T} 
\]

@@

The first step consists in turning $\V{T}$ into a predicate
$\Bhab{\V{P}}{\Mu{\V{D}} \To \Set}$, hence introducing a dependency on the
descriptions. Consequently, we strengthen the algebra $\descop{\V{D}}{\V{T}}
\To \V{T}$ by a similar process. We start with turning the arrow into a
dependent arrow, hence offering $\Bhab{\V{xs}}{\descop{\V{D}}{(\Mu{\V{D}})}}$ to
further analysis.

Then, we have to translate the notion of algebra in our dependent
setting. This is achieved by the predicate $\All{\V{D}}{\V{X}}{\V{P}}{\V{x}}$
(Fig.~\ref{fig:all-predicates}) that, intuitively, states that
$\Bhab{\V{P}}{\V{X} \To \Set}$ holds everywhere in $\V{x}$. Based on that
knowledge, we ought to be able to prove that $\V{P}$ holds for
$\Con{\V{xs}}$. We close our elimination principle by taking an argument
$x$ in $\Mu{\V{D}}$ and claiming that the motive $\V{P}$ is realised on $\V{x}$.

\[
\begin{array}{lcll}
\F{induction} & : & \multicolumn{2}{l}{\PITEL{\V{D}}{\Desc}}                   \\
              &   & \multicolumn{2}{l}{\PITEL{\V{x}}{\Mu{\V{D}}}}                   \\
              &   & \multicolumn{2}{l}{\PITEL{\V{P}}{\Mu{\V{D}} \To \Set}}         \\
              &   & \PI{\V{m}}{& \PI{\V{xs}}{\descop{\V{D}}{(\Mu{\V{D}})}}              \\
              &   &           & \All{\V{D}}{(\Mu{\V{D}})}{\V{P}}{\V{xs}} \To \V{P} (\Con{\V{xs}})} \\
              &   & \V{P} \V{x}
\end{array}
\]

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\All{}{}{}{} : & \PITEL{\V{D}}{\Desc}
                 \PITEL{\V{X}}{\Set}
                 \PITEL{\V{P}}{\V{X} \To \Set} \\
               & \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                 \Set 
\end{array} \\
\All{\DId}{\V{X}}{\V{P}}{\V{x}} \mapsto \V{P}\: \V{x} \\
\All{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{z}} \mapsto \Unit \\
\All{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\pair{\V{d}}{\V{d'}}{}} \mapsto \TIMES{\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}}{\All{\V{D'}}{\V{X}}{\V{P}}{\V{d'}}} \\
\All{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} \mapsto \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{b}} \\
\All{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{f}} \mapsto \PI{\V{\V{s}}}{\V{S}} \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{(\V{f}\: \V{s})}
}
&
%%
\stk{
\begin{array}{@{}ll}
\all{}{}{}{}{} : & \PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X}}{\Set}
                   \PITEL{\V{P}}{\V{X} \To \Set} \\
                 & \PITEL{\V{R}}{\PI{\V{x}}{\V{X}} \V{P}\: \V{x}}
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{\V{P}}{\V{xs}} 
\end{array} \\
\all{\DId}{\V{X}}{\V{P}}{\V{R}}{\V{x}} = \V{R}\: \V{x} \\
\all{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{R}}{\V{z}} = \Void \\
\all{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{d}}{\V{d'}}{}} = \pair{\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{d}}}{\all{\V{D'}}{\V{X}}{\V{P}}{\V{R}}{\V{d'}}}{} \\
\all{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} = \all{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{R}}{\V{b}} \\
\all{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\V{f}} = \LAM{\V{s}} \all{(\V{T}\: \V{s})}{\V{X}}{\V{P}}{\V{R}}{(\V{f}\: \V{s})}
\end{array}
}
\]

@@

\[
\Rule{\forall \V{e_i} \in \V{\vec{e}}, \propag{\push{\V{e_i}}
                                            {\descop{\V{f}\: \V{i}}{(\Mu{(\DSigma{(\Enum\: \V{b})}{\V{f}})})}}}
                                      {\V{e'_i}}}
     {\propag{\push{\V{C}\: \V{\vec{e}}}{\Mu{(\DSigma{(\Enum\: \V{b})}{\V{f}})}}}
             {\Con{(\etag{C}, \V{\widetilde{e'}})}}}\;\etag{t} \in \V{b}
\]
